\textbf{DiracDec C++/Coq Implementation\cite{ExampleCitation}}

### 2024-11-25 main.tex
CoC typings

\begin{figure}[h]
    \begin{align*}
        & \text{(R-FLATTEN)} &&
        \textcolor{gray}{\text{A special rule to flatten all AC symbols within one call.}} \\
        & \text{(R-ADDSID)} &&
        \ADDS(a) \to a \\
        & \text{(R-ADDS0)} &&
        \ADDS((seq:\_ \_)\ \ZERO) \to \ADDS(seq) \\
        & && \textcolor{gray}{\text{This rule removes all $\ZERO$ occurances and keeps the order of remaining subterms.}} \\
        & \text{(R-MLTSID)} &&
        \MLTS(a) \to a \\
        & \text{(R-MLTS0)} &&
        \MLTS((seq: \_ \_)\ \ZERO) \to \ZERO \\
        & \text{(R-MLTS1)} &&
        \MLTS((seq: \_ \_)\ \ONE) \to \MLTS(seq) \\
        & && \textcolor{gray}{\text{Similar to (R-ADDS0).}} \\
        & \text{(R-MLTS2)} &&
        \MLTS((seq1: \_ \_ \_)\ \ADDS(a_1\ a_2 \cdots a_n)\ (seq2: \_ \_ \_)) \\
        & && \to \ADDS(\MLTS(seq1\ a_1\ seq2)\ \MLTS(seq1\ a_2\ seq2) \cdots \MLTS(seq1\ a_n\ seq2)) \\
        & && \textcolor{gray}{\text{This rule matches the first $\ADDS$ subterm in the list. $seq1$ or $seq2$ can be empty, but not both.}} \\
        & \text{(R-CONJ0)} &&
        \CONJ(\ZERO) \to \ZERO \\
        & \text{(R-CONJ1)} &&
        \CONJ(\ONE) \to \ONE \\
        & \text{(R-CONJ2)} &&
        \CONJ(\ADDS(a_1\ a_2\ \cdots a_n)) \to \ADDS(\CONJ(a_1)\ \CONJ(a_2) \cdots \CONJ(a_n)) \\
        & \text{(R-CONJ3)} &&
        \CONJ(\MLTS(a_1\ a_2\ \cdots a_n)) \to \MLTS(\CONJ(a_1)\ \CONJ(a_2) \cdots \CONJ(a_n)) \\
        & \text{(R-CONJ4)} &&
        \CONJ(\CONJ(a)) \to a
    \end{align*}
\end{figure}

@@

% \section{About Typing}
% I plan to implement a typing system in C++, replicating the one in Coq.

>>>>

\begin{figure}[h]
    \begin{align*}
        & \text{(R-FLATTEN)} &&
        \textcolor{gray}{\text{A special rule to flatten all AC symbols within one call.}} \\
        & \text{(R-ADDSID)} &&
        \ADDS(a) \to a \\
        & \text{(R-ADDS0)} &&
        \ADDS((seq:\_ \_)\ \ZERO) \to \ADDS(seq) \\
        & && \textcolor{gray}{\text{This rule removes all $\ZERO$ occurances and keeps the order of remaining subterms.}} \\
        & \text{(R-MLTSID)} &&
        \MLTS(a) \to a \\
        & \text{(R-MLTS0)} &&
        \MLTS((seq: \_ \_)\ \ZERO) \to \ZERO \\
        & \text{(R-MLTS1)} &&
        \MLTS((seq: \_ \_)\ \ONE) \to \MLTS(seq) \\
        & && \textcolor{gray}{\text{Similar to (R-ADDS0).}} \\
        & \text{(R-MLTS2)} &&
        \MLTS((seq1: \_ \_ \_)\ \ADDS(a_1\ a_2 \cdots a_n)\ (seq2: \_ \_ \_)) \\
        & && \to \ADDS(\MLTS(seq1\ a_1\ seq2)\ \MLTS(seq1\ a_2\ seq2) \cdots \MLTS(seq1\ a_n\ seq2)) \\
        & && \textcolor{gray}{\text{This rule matches the first $\ADDS$ subterm in the list. $seq1$ or $seq2$ can be empty, but not both.}} \\
        & \text{(R-CONJ0)} &&
        \CONJ(\ZERO) \to \ZERO \\
        & \text{(R-CONJ1)} &&
        \CONJ(\ONE) \to \ONE \\
        & \text{(R-CONJ2)} &&
        \CONJ(\ADDS(a_1\ a_2\ \cdots a_n)) \to \ADDS(\CONJ(a_1)\ \CONJ(a_2) \cdots \CONJ(a_n)) \\
        & \text{(R-CONJ3)} &&
        \CONJ(\MLTS(a_1\ a_2\ \cdots a_n)) \to \MLTS(\CONJ(a_1)\ \CONJ(a_2) \cdots \CONJ(a_n)) \\
        & \text{(R-CONJ4)} &&
        \CONJ(\CONJ(a)) \to a
    \end{align*}
    \caption{The rewriting rules for scalars.}
\end{figure}

@@

\clearpage
\section{Typing System}
The typing system modifies the one for Calculus of Construction, adding special symbols for Dirac notations.
According to the CoC typing rules, the environment and context can be modeled by as two stacks. It is obvious that popping assumptions/definitions from any context or from a environment with empty context will not destroy the well-formedness.

One important reason for using dependent type is that, the types of Dirac notatinos (i.e., ket, bra and operators) depends on the basis.

\newcommand{\reduce}{\triangleright}

\newcommand{\SORT}{\mathsf{Sort}}
\newcommand{\WF}{\mathcal{WF}}
\newcommand{\Type}{\mathsf{Type}}
\newcommand{\BaseS}{\mathsf{Base}}

\newcommand{\SType}{\mathcal{S}}
\newcommand{\KType}{\mathcal{K}}
\newcommand{\BType}{\mathcal{B}}
\newcommand{\OType}{\mathcal{O}}

\newcommand{\ZEROK}{\mathbf{0}_\mathcal{K}}
\newcommand{\ZEROB}{\mathbf{0}_\mathcal{B}}
\newcommand{\ZEROO}{\mathbf{0}_\mathcal{O}}
\newcommand{\ONEO}{\mathbf{1}_\mathcal{O}}

\begin{figure}[h]
    \begin{gather*}
        \textbf{W-Empty} \qquad
        \frac{}{\WF([])[]}\\
        \\
        \textbf{W-Local-Assum} \qquad
        \frac{E[\Gamma]\vdash T:s \qquad s \in \mathcal{S} \qquad x \notin \Gamma}{\WF(E)[\Gamma::(x:T)]}
        \qquad \qquad
        \textbf{W-Local-Def} \qquad
        \frac{E[\Gamma]\vdash t : T \qquad x \notin \Gamma}{\WF(E)[\Gamma ::(x:=t:T)]} \\
        \\
        \textbf{W-Global-Assum} \qquad
        \frac{E[]\vdash T:s \qquad s \in \mathcal{S} \qquad c \notin E}{\WF(E; c:T)[]}
        \qquad \qquad
        \textbf{W-Global-Def} \qquad
        \frac{E[]\vdash t:T \qquad c \notin E}{\WF(E; c:=t:T)[]} \\
        \\
        \textbf{Ax-Type} \qquad
        \frac{\WF(E)[\Gamma]}{E[\Gamma]\vdash \Type(i):\Type(i+1)} \\
        \\
        \textbf{Var} \qquad
        \frac{\WF(E)[\Gamma] \qquad (x:T) \in \Gamma \text{ or } (x := t : T) \in \Gamma \text{ for some $t$}}{E[\Gamma] \vdash x : T} \\
        \\
        \textbf{Const} \qquad
        \frac{\WF(E)[\Gamma] \qquad (c:T) \in E \text{ or } (c := t : T) \in E \text{ for some $t$}}{E[\Gamma] \vdash c : T} \\
        \\
        \textbf{Prod-Type} \qquad
        \frac{E[\Gamma] \vdash T:\Type(i)\qquad E[\Gamma::(x:T)]\vdash U:\Type(i)}{E[\Gamma] \vdash \forall x: T, U : \Type(i)} \\
        \\
        \textbf{Lam} \qquad
        \frac{E[\Gamma] \vdash \forall x:T, U : s \qquad E[\Gamma :: (x:T)]\vdash t : U}{E[\Gamma] \vdash \lambda x:T . t : \forall x : T, U} \\
        \\
        \textbf{App} \qquad
        \frac{E[\Gamma] \vdash t:\forall x:U, T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T\{x/u\}} \\
        \\
        \textbf{Subtyping} \qquad
        \frac{E[\Gamma] \vdash U : s \qquad E[\Gamma] \vdash t : T \qquad E[\Gamma] \vdash T \leq_{\beta\delta\eta} U}{E[\Gamma]\vdash t:U}
    \end{gather*}
    \caption{
    The typing rules for CoC. 
    We only consider computational types, removing the sorts of $\mathsf{Set}$ and $\mathsf{Prop}$.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \frac{E[\Gamma]\vdash t =_{\beta\delta\eta} u}{E[\Gamma]\vdash t \leq_{\beta\delta\eta} u} 
        \qquad
        \frac{i \leq j}{E[\Gamma] \vdash \Type(i) \leq_{\beta\delta\eta} \Type(j)}
        \qquad
        \frac{E[\Gamma] \vdash T =_{\beta\delta\eta} U \qquad E[\Gamma :: (x:T)]\vdash T' \leq_{\beta\delta\eta} U'}{E[\Gamma] \vdash \forall x : T, T' \leq_{\beta\delta\eta} \forall x : U, U'}
    \end{gather*}
    \caption{The rules to decide subtyping.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Type-Base} \qquad
        \frac{\WF(E)[\Gamma]}{E[\Gamma]\vdash \BaseS : \Type(1)} 
        \qquad \qquad
        \textbf{Type-Scalar} \qquad
        \frac{\WF(E)[\Gamma]}{E[\Gamma]\vdash \SType : \Type(1)} \\
        \\
        \textbf{Type-Ket} \qquad
        \frac{E[\Gamma]\vdash \sigma : \BaseS}{E[\Gamma]\vdash \KType(\sigma) : \Type(1)}
        \qquad \qquad
        \textbf{Type-Bra} \qquad
        \frac{E[\Gamma]\vdash \sigma : \BaseS}{E[\Gamma]\vdash \BType(\sigma) : \Type(1)} \\
        \\
        \textbf{Type-Opt} \qquad
        \frac{E[\Gamma]\vdash \sigma : \BaseS \qquad E[\Gamma]\vdash \tau : \BaseS}{E[\Gamma]\vdash \OType(\sigma, \tau) : \Type(1)}
    \end{gather*}
    \caption{Typing rules for Dirac notation types.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Sca-0} \qquad
        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash 0 : \SType}
        \qquad \qquad
        \textbf{Sca-1} \qquad
        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash 1 : \SType}
        \qquad \qquad
        \textbf{Sca-Delta} \qquad
        \frac{            
            \begin{aligned}
                & E[\Gamma]\vdash \sigma : \BaseS \\
                & E[\Gamma]\vdash s : \sigma \qquad E[\Gamma]\vdash t : \sigma
            \end{aligned}} {E[\Gamma] \vdash \delta_{s, t} : \SType} \\
        \\
        \textbf{Sca-Add} \qquad
        \frac{E[\Gamma]\vdash a : \SType \qquad E[\Gamma]\vdash b : \SType}{E[\Gamma]\vdash a + b : \SType}
        \qquad \qquad
        \textbf{Sca-Mul} \qquad
        \frac{E[\Gamma]\vdash a : \SType \qquad E[\Gamma]\vdash b : \SType}{E[\Gamma]\vdash a \times b : \SType} \\
        \\
        \textbf{Sca-Conj} \qquad
        \frac{E[\Gamma] \vdash a : \SType}{E[\Gamma] \vdash a^*:\SType}
        \qquad \qquad
        \textbf{Sca-Dot} \qquad
        \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
    \end{gather*}
    \caption{Scalar typing rules.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Base-Def} \qquad
        \frac{\WF(E)[\Gamma] \qquad \text{for all $i$, } a_i \notin \Gamma \text{ and } a_i \notin E}
        {E[\Gamma]\vdash \{a_1, a_2, \cdots, a_n\} : \BaseS} \\
        \\
        \textbf{Base-Element} \qquad
        \frac{\WF(E)[\Gamma] \qquad c \in \{a_1, a_2, \cdots a_n\}}{E[\Gamma]\vdash c : \{a_1, a_2, \cdots, a_n\}} \\
        \\
        \textbf{Base-Prod} \qquad
        \frac{E[\Gamma]\vdash \sigma : \BaseS \qquad E[\Gamma]\vdash \tau : \BaseS}{E[\Gamma]\vdash \sigma \times \tau : \BaseS} \\
        \\
        \textbf{Pair-Base} \qquad
        \frac{
            \begin{aligned}
                & E[\Gamma]\vdash \sigma : \BaseS \qquad E[\Gamma]\vdash s : \sigma \\
                & E[\Gamma]\vdash \tau : \BaseS \qquad E[\Gamma]\vdash t : \tau
            \end{aligned}} {E[\Gamma]\vdash (s, t) : \sigma \times \tau} 
        \textbf{}
    \end{gather*}
    \caption{
    Typing rules for bases. 
    Here the definitions $\{a_1, a_2, \cdots a_n\}$ of different order elements are equivalent.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Ket-0} \qquad
        \frac{E[\Gamma] \vdash \sigma : \BaseS}{E[\Gamma] \vdash \ZEROK(\sigma) : \KType(\sigma)} 
        \qquad \qquad
        \textbf{Ket-Base} \qquad
        \frac{E[\Gamma] \vdash \sigma : \BaseS \qquad E[\Gamma]\vdash t : \sigma}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)} \\
        \\
        \textbf{Ket-Adj} \qquad
        \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B^\dagger : \KType(\sigma)} 
        \qquad \qquad
        \textbf{Ket-Sca} \qquad
        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash a.K : \KType(\sigma)} \\
        \\
        \textbf{Ket-Add} \qquad
        \frac{E[\Gamma] \vdash K_1 : \KType(\sigma) \qquad E[\Gamma] \vdash K_2 : \KType(\sigma)}{E[\Gamma] \vdash K_1 + K_2 : \KType(\sigma)} \\
        \\
        \textbf{Ket-Dot} \qquad
        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash K : \KType(\tau)}{E[\Gamma] \vdash O \cdot K : \KType(\sigma)} \\
        \\
        \textbf{Ket-Tsr} \qquad
        \frac{E[\Gamma] \vdash K_1 : \KType(\sigma) \qquad E[\Gamma] \vdash K_2 : \KType(\tau)} {E[\Gamma] \vdash K_1 \otimes K_2 : \KType(\sigma \times \tau)}
    \end{gather*}
    \caption{Ket typing rules.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Bra-0} \qquad
        \frac{E[\Gamma] \vdash \sigma : \BaseS}{E[\Gamma] \vdash \ZEROB(\sigma) : \BType(\sigma)} 
        \qquad \qquad
        \textbf{Bra-Base} \qquad
        \frac{E[\Gamma] \vdash \sigma : \BaseS \qquad E[\Gamma]\vdash t : \sigma}{E[\Gamma] \vdash \bra{t} : \BType(\sigma)} \\
        \\
        \textbf{Bra-Adj} \qquad
        \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K^\dagger : \BType(\sigma)} 
        \qquad \qquad
        \textbf{Bra-Sca} \qquad
        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash a.B : \BType(\sigma)} \\
        \\
        \textbf{Bra-Add} \qquad
        \frac{E[\Gamma] \vdash B_1 : \BType(\sigma) \qquad E[\Gamma] \vdash B_2 : \BType(\sigma)}{E[\Gamma] \vdash B_1 + B_2 : \BType(\sigma)} \\
        \\
        \textbf{Bra-Dot} \qquad
        \frac{E[\Gamma] \vdash B : \KType(\sigma) \qquad E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash B \cdot O : \BType(\tau)} \\
        \\
        \textbf{Bra-Tsr} \qquad
        \frac{E[\Gamma] \vdash B_1 : \BType(\sigma) \qquad E[\Gamma] \vdash B_2 : \BType(\tau)} {E[\Gamma] \vdash B_1 \otimes B_2 : \BType(\sigma \times \tau)}
    \end{gather*}
    \caption{Bra typing rules.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Opt-0} \qquad
        \frac{E[\Gamma] \vdash \sigma : \BaseS \qquad E[\Gamma] \vdash \tau : \BaseS}{E[\Gamma] \vdash \ZEROO(\sigma, \tau) : \OType(\sigma, \tau)} 
        \qquad \qquad
        \textbf{Opt-1} \qquad
        \frac{E[\Gamma] \vdash \sigma : \BaseS}{E[\Gamma] \vdash \ONEO(\sigma) : \OType(\sigma, \sigma)} \\
        \\
        \textbf{Opt-Adj} \qquad
        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O^\dagger : \OType(\tau, \sigma)} 
        \qquad \qquad
        \textbf{Opt-Sca} \qquad
        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash a.O : \OType(\sigma, \tau)} \\
        \\
        \textbf{Opt-Add} \qquad
        \frac{E[\Gamma] \vdash O_1 : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash O_2 : \OType(\sigma, \tau)}{E[\Gamma] \vdash O_1 + O_2 : \OType(\sigma, \tau)} \\
        \\
        \textbf{Opt-Outer} \qquad
        \frac{E[\Gamma]\vdash K : \KType(\sigma) \qquad E[\Gamma]\vdash B : \BType(\tau)}{E[\Gamma]\vdash K \cdot B : \OType(\sigma, \tau)} \\
        \\
        \textbf{Opt-Dot} \qquad
        \frac{E[\Gamma] \vdash O_1 : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash O_2 : \OType(\tau, \rho)}{E[\Gamma] \vdash O_1 \cdot O_2 : \OType(\sigma, \rho)} \\
        \\
        \textbf{Opt-Tsr} \qquad
        \frac{E[\Gamma] \vdash O_1 : \OType(\sigma_1, \tau_1) \qquad E[\Gamma] \vdash O_2 : \OType(\sigma_2, \tau_2)} {E[\Gamma] \vdash O_1 \otimes O_2 : \OType(\sigma_1 \times \sigma_2, \tau_1 \times \tau_2)}
    \end{gather*}
    \caption{Operator typing rules.}
\end{figure}

\clearpage
\section{Conversions and Reductions}
\begin{figure}[h]
    \begin{gather*}
        \textbf{Alpha} \qquad \text{(bound variable renaming)}
        \qquad \qquad
        \textbf{Beta} \qquad
        \frac{}{E[\Gamma] \vdash ((\lambda x:T.t)\ u) \reduce_\beta t\{x/u\}} \\
        \\
        \textbf{Delta-Local} \qquad
        \frac{\WF(E)[\Gamma] \qquad (x:=t:T) \in \Gamma}{E[\Gamma] \vdash x \reduce_\Delta t} 
        \qquad
        \textbf{Delta-Global} \qquad
        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c \reduce_\delta t} \\
        \\
        \textbf{Eta} \qquad
        \frac{E[\Gamma]\vdash t : \forall x : T , U}{E[\Gamma] \vdash t =_\eta \lambda x : T . (t\ x)}
    \end{gather*}
    \caption{Conversions and reductions in CoC.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \cdots
    \end{gather*}
    \caption{Dirac notation reduction rules}
\end{figure}

About big operators: maybe we should use the "sum operators"?

\clearpage
